---
title: "VO2max"
author: "Sara G"
date: "21/05/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## step 3 reading in the data and checking structure, head and tail

```{r packages, echo= FALSE}
library(tidyverse) # include all required packages at the start
library(broom)
library(dplyr)
```

## reading in the files

```{r excel} 
df <- readxl::read_excel("data/vo2-max_data.xlsx")
```

### Structure

```{r str}
str(df)
```

```{r head}
head(df)
```

```{r tail}
tail(df)
```

## Data transformation

changing the class of the gender factor

```{r transform}
df$gender <- as.factor(df$gender)
str(df)
```

## Step 5 Exploratory data analysis

```{r visualisation age}
## 1. age and VO2max
ggplot(df, aes(x = age, y = vo2_max)) +
  geom_point(alpha = 0.5, colour = "darkgreen") +
  geom_smooth(method = "lm")
```

```{r visualisation weight}
## 1. age and VO2max
ggplot(df, aes(x = weight, y = vo2_max)) +
  geom_point(alpha = 0.5, colour = "purple") +
  geom_smooth(method = "lm")
```

```{r visualisation hear_rate}
## 1. age and VO2max
ggplot(df, aes(x = heart_rate, y = vo2_max)) +
  geom_point(alpha = 0.5, colour = "brown") +
  geom_smooth(method = "lm")
```

```{r visualisation gender}
## 1. age and VO2max
ggplot(df, aes(x = gender, y = vo2_max)) +
  geom_point(alpha = 0.5, colour = "red") +
  geom_smooth(method = "lm")
```

## weight and age tend to have a negative correlation with VO2max. The correlation is moderate with weight and mild with age. There doesn't seem to be much correlation with heart rate. and there is no relationship at all with gender.

## Step 5.2

```{r age & VO2max with separate gender}
ggplot(df, aes(x = age, y = vo2_max, colour = gender)) +
  geom_point() +
  geom_smooth(method = "lm")
```

There is a minor relationship between males and their age and VO2max, but not with females


```{r weight & VO2max with separate gender}
ggplot(df, aes(x = weight, y = vo2_max, colour = gender)) +
  geom_point() +
  geom_smooth(method = "lm")
```

There seems to be a moderate relationship between weight & age and Vo2max for both male and females

```{r heart_rate & VO2max with separate gender}
ggplot(df, aes(x = heart_rate, y = vo2_max, colour = gender)) +
  geom_point() +
  geom_smooth(method = "lm")
```

There seems to be minimal to no relationship between heart_rate and VO2max when accounting for gender. 

```{r box plot}
ggplot(data = df, aes(x = gender, y = vo2_max)) +
  geom_boxplot(aes(fill = gender))
 
```

## Step 5.3 Multicollinearity

```{r muticollinearity visualisation}
pairs(formula = ~ age + weight + heart_rate, data = df)
```

There doesn't seem to be much relationship between any of the explanatory variables


## Step 6 Mutiple linear regression

```{r Multiple linear regression}
library(broom)
fit <- lm(vo2_max ~ age + weight + heart_rate + gender, data = df)
tidy(fit, conf.int = TRUE)
```

```{r summary MLR}
summary(fit)
```

# intercept: when age, weight, heart_rate (all the continuous variables) = 0 and when gender = female, then the estimated VO2max is 87.8 ml/kg/min. 

#age: if we increase age by one year and hold the other variables constant (fixed), then we can expect the VO2max to decrease by 0.165. with a variability of between -0.2 to -0.04 if we chose another sample. This makes sense

#weight: if we increase weight by one kg and hold the other variables constant, then we can expect the VO2max to decrease by 0.3846. with a variability of between -0.471 to -0.2985 if we chose another sample. This makes sense

#heart_rate: if we increase heart_rate by one bpm and hold the other variables constant, then we can expect the VO2max to decrease by 0.118ml/kg/min with a variability of between -0.181 to -0.0541 if we chose another sample. This makes sense

#gender - male: males have a VO2max 13.2ml/kg/min higher on average, than females, when all the other variables are held constant. 

## Step 7 Independence

```{r independence}
car::durbinWatsonTest(fit)
```
## OUr DW value is very close to 2 so we can assume that our indepdence of observation has been met

## Step 8 Outliers

```{r outliers}
std_res <- rstandard(fit)
points <- 1:length(std_res)
res_labels <- if_else(abs(std_res) >= 2.5, paste(points), "")

ggplot(data = NULL, aes(x = points, y = std_res)) +
  geom_point() +
  geom_text(aes(label = res_labels), nudge_y = 0.3)+
  ylim(c(-4, 4)) +
  geom_hline(yintercept = c(-3,3), colour = "red", linetype = "dashed")
```

There looks like there is 3 outliers



## Step 9 Levarage points

```{r leverage}
hats <- hatvalues(fit)
hat_labels <-  if_else(hats >= 0.15, paste(points), "")


ggplot(data = NULL, aes(x = points, y = hats)) +
  geom_point() +
  geom_text(label = hat_labels, nudge_y = 0.005)

```

## Step 10 Influential points

```{r cook}
cook <- cooks.distance(fit)
cook_labels <- if_else(cook >= 0.075, paste(points), "")

ggplot(data = NULL, aes(x = points, y = cook)) +
  geom_point() +
  geom_text(aes(label = cook_labels), nudge_y = 0.005)
```

## filtering out outliers, by creating a new df

```{r outliers_new}
outliers <- c(25,28,32,35,41,73) # store participant IDs to remove injobect 'outliers'
filtered_df <- df %>%
  filter(!subject_id %in% outliers) # filter if subject_id is not (!) %in% outliers
```

## re-run the linear regression with filtered_df

```{r re_run MLR}
fit2 <- lm(vo2_max ~ age + weight + heart_rate + gender, data = filtered_df)
tidy(fit2, conf.int = TRUE)
```

```{r summary MLR2}
summary(fit2)
```
## the R squared value has increased from 0.577 to 0.767. therefore we can say that our coefficient of determination  us that 76.7% of the variabnce in VO2max can be explaied by the variance in these continuous variables above.  

## by removing these outliers the slope and intercept have all increased significantly for weight and gender in particularly.

## we should check our outliers/influential data points to make sure that they are not due to measurement error
## if they are not due to mearusement error we might consider transforming the variables and re-running the model, or choosing a different model type that is more robust to these points. REmoving real data should be a last option. 

## Step 11. Homoscedasticity

```{r homoscedasticity}
res <- residuals(fit) # all the distances of the observed value from the line of best fit
fitted <- predict(fit) # the predicted value line (x)

ggplot(data = NULL, aes(x = fitted, y = res)) +
  geom_point(colour = "purple") +
  geom_hline(yintercept = 0, colour = "red", linetype = "dashed")


```

# there is no clear pattern, therefore we can assume homoscedasticity

## Step 12 Normality

```{r normality}
ggplot(data = NULL, aes(x = res)) +
  geom_histogram (colour = "black", fill = "purple", binwidth = 3)
```

# the seem to be relatively normally distributed

```{r q-q plot}
ggplot(data = NULL, aes(sample = res)) +
  stat_qq() + stat_qq_line()
```
## There is signs of linearity here so we can assum that we have a normal distribution and we haven't failed the assumption of normality

# Step 13 Multicollinearity

```{r variance inflation factor}
car::vif(fit)
```

## Our factors are close to one, there is not much multicollinearity 
```{r Standard error}
sqrt(car::vif(fit))
```
## The SE is minimal, the assumption of multicollinearity is assumed. 

## Step 14 Linearity

```{r linearity}
car::avPlots(fit)
```
Linearity is assumed

## Step 15 Interpretation

```{r VO2max Equation}
## VO2max = 87.83030 + (age * -0.16514) + (-0.38457 * weight) + (-0.11791 * HR) + (13.2 * gender)
87.83030 + (31 * -0.16514) + (-0.38457 * 68) + (-0.11791 * 140) + (13.2 * 0)
```

## Limitations
## Some of the limitations include there are some outliers that may be influencing our data and these could be due to measurement error. We may therefore need to try to transform our data via e.g. log transformation
## We might consider trying non-linear models that are more robust with our data. 
## we could also consider looking at the observed vs predicted to see what the error is in our data. 


```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
